---
title: "Linear Response Plateau (LRP)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Response Plateau (LRP)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 120)
```

```{r setup}
library(DimExp)
```

## Theory

In a uniformity trial, adjacent basic experimental units (BEU) are grouped into
plots of increasing size `X`, and the coefficient of variation `CV(X)` is
computed among plots of each size. The CV falls as plots get larger, but the
gain shrinks: past a certain size, extra area buys almost no precision. The
Linear Response Plateau model formalizes that as two joined straight lines,

$$
CV_{(X)} =
\begin{cases}
a + bX + \varepsilon, & \text{if } X \le X_o \\
p + \varepsilon,      & \text{if } X > X_o
\end{cases}
$$

The first segment falls with slope $b < 0$; the second is flat at the plateau
$p$. The join between them is the estimate of interest: $X_o$ is the optimal
plot size and $p = CV_{Xo}$ is the CV expected at that size. From the model,

$$
X_o = \frac{p - a}{b}, \qquad CV_{Xo} = a + b X_o
$$

### How the breakpoint is estimated

Fitting a broken line is awkward for general-purpose nonlinear least squares:
the model has a kink, so `nls()` depends heavily on the starting values and can
settle on a local minimum or fail to converge without saying so clearly.

`fit_lrp()` avoids this by **profiling** the breakpoint. It walks a fine grid of
candidate values for $X_o$; at each candidate the model becomes linear, so the
coefficients come from ordinary least squares. The candidate with the smallest
residual sum of squares wins. There are no starting values, no local minima, and
no convergence failures.

### References

Original method: Paranaíba, P. F., Ferreira, D. F. & Morais, A. R. (2009).
Tamanho ótimo de parcelas experimentais: proposição de métodos de estimação.
*Revista Brasileira de Biometria*, 27(2), 255-268.

Validation used here: Cargnelutti Filho, A., Loro, M. V., Ortiz, V. M. &
Andretta, J. A. (2025). Determinação do tamanho de parcela para avaliar a massa
de parte aérea de grão-de-bico. *Revista Vivências*, 21(43), 499-513.

## Data

The examples use the chickpea uniformity trials of the validation article: a
6 × 6 grid of 1 m² BEU per trial, grouped into 15 plot shapes. `X` is the plot
size in BEU and `CV1`--`CV3` are the CVs of trials 1 to 3.

```{r data}
X   <- c(1, 2, 2, 3, 3, 4, 6, 6, 6, 6, 9, 12, 12, 18, 18)
CV1 <- c(30.40, 19.51, 23.72, 12.89, 21.32, 16.69, 6.71, 10.75,
         17.58, 14.94, 11.93, 3.18, 8.63, 4.25, 11.41)
CV2 <- c(34.79, 29.44, 26.45, 26.35, 31.74, 25.01, 21.44, 22.86,
         28.93, 16.17, 27.29, 19.89, 16.85, 26.53, 14.70)
CV3 <- c(31.12, 20.87, 29.09, 20.02, 28.04, 19.97, 10.74, 19.28,
         18.18, 26.46, 17.41, 11.51, 17.96, 9.18, 19.17)
```

Note that `X` repeats: several shapes share the same size (four different
shapes of 6 BEU, for instance). Supply one CV per shape, never the mean per
size, or the fit loses the within-size variability it needs.

## Basic use

```{r fit}
fit <- fit_lrp(x = X, cv = CV1)
fit
```

$X_o = 7.44$ m² with $CV_{Xo} = 7.88\%$, reproducing the published values for
trial 1.

`summary()` splits the model coefficients from the fit statistics:

```{r summary}
summary(fit)
```

Everything is also available directly:

```{r access}
fit$coefficients
fit$parameters["Breakpoint"]
round(fit$residuals[1:5], 3)
```

`predict()` evaluates the fitted model at any plot size, which is handy to check
the plateau behaviour:

```{r predict}
predict(fit, newx = c(1, 5, 7.44, 15))
```

The last two values are identical: past the breakpoint the model is flat.

## Plot

The `plot()` method draws the article-style figure. Title and styling go here,
not in `fit_lrp()`:

```{r plot}
plot(fit, title = "Trial 1 (chickpea)")
```

For Portuguese-language figures, switch the decimal mark and the conditional
word:

```{r plot-ptbr}
plot(fit, title = "Ensaio 1", decimal_mark = ",", cond_word = "se")
```

Other useful arguments: `annotate_model = FALSE` drops the equation block,
`base_size` / `label_size` / `title_size` scale the text, `theme` accepts any
ggplot2 theme, and `xlab` / `ylab` relabel the axes.

To export, set `save = TRUE`. TIFF is written with LZW compression; PDF and EPS
are vector formats, usually the better choice for line art like this.

```{r save, eval = FALSE}
plot(fit, title = "Trial 1",
     save = TRUE, file = "trial1.tiff", format = "tiff",
     dpi = 300, width = 18, height = 12, units = "cm")
```

## Several trials at once

Pass a data frame plus the column names. One model is fitted per trial:

```{r multi}
trials <- rbind(
  data.frame(x = X, cv = CV1, trial = "Trial 1"),
  data.frame(x = X, cv = CV2, trial = "Trial 2"),
  data.frame(x = X, cv = CV3, trial = "Trial 3")
)

res <- fit_lrp(trials, x = "x", cv = "cv", trial = "trial")
res
```

The three breakpoints (7.44, 6.56, 7.57) match the published table, and their
mean, 7.19 m², is the article's overall recommendation.

The individual fits stay accessible, so any single trial can be inspected or
plotted as usual:

```{r multi-access}
res$fits[["Trial 2"]]
```

`plot()` on the multi-trial object arranges the panels in a grid of at most six
(3 rows × 2 columns), paginating into separate figures beyond that. It needs
the **patchwork** package.

```{r multi-plot, eval = FALSE}
plot(res, decimal_mark = ",", cond_word = "se", label_size = 3)
```

Column names default to `"x"`, `"cv"` and `"trial"`, so `fit_lrp(trials)` alone
works when the data frame already uses those names. Missing columns produce a
clear error rather than a cryptic one:

```{r error, error = TRUE}
fit_lrp(trials, x = "x", cv = "CV", trial = "trial")
```

## Fine-tuning

**`search_range`** restricts where the breakpoint is searched, which helps when
an outlier in the transition zone pulls the estimate somewhere implausible.
It must fall inside the data range, and a warning is issued if the breakpoint
lands on the edge of the range you gave.

```{r search-range}
fit_lrp(X, CV1, search_range = c(5, 12))$parameters["Breakpoint"]
```

**`step`** is the grid resolution, 0.001 by default. Coarser grids are faster
and slightly less precise:

```{r step}
fit_lrp(X, CV1, step = 0.01)$parameters["Breakpoint"]
```

**`method`** chooses how the linear coefficients are estimated at each candidate
breakpoint. `"segment"` (default) uses only the observations with `x <= X0`,
reproducing the published procedure. `"ramp"` uses every observation on the
basis `pmin(x, X0)`, the standard least-squares LRP; the plateau points then
also inform the descending line, which can shift the optimum.

```{r method}
fit_lrp(X, CV1, method = "ramp")$parameters["Breakpoint"]
```

## Warnings worth heeding

Two situations trigger a warning, and both mean the estimate deserves a second
look rather than direct use:

- **Non-negative slope**: the CV is not decreasing with plot size, so the
  "descent then plateau" reading does not hold for these data.
- **Breakpoint at the edge of the search range**: the trial may not span enough
  plot sizes to bracket the true optimum, so the CV may still be falling at the
  largest size evaluated.

## Where to go next

`vignette("qrp")` and `vignette("mcm")` cover the other two CV-based methods,
which take the same inputs and usually give a larger and a smaller optimum
respectively. `vignette("replicates")` shows how $CV_{Xo}$ feeds into the number
of replications.
